The very dependent recursive structure of iterated parametricity in indexed form

### 2025-12-29 paper/paper.tex
paper: minor change, whitespace
paper: strip a block of equations

\newcommand{\Prop}{\textsf{Prop}}
\newcommand{\MySet}{\textsf{Set}}

Parametricity, like realisability, reducibility, and logical relations, was most commonly handled in an \emph{indexed} form, that is, using a type-theoretic language, considering predicates of type $A \rightarrow \Prop$, or families of type $A \rightarrow \MySet$, or relations of type $A \times A \rightarrow \Prop$, or relevant relations -- sometimes called \emph{correspondences} -- of type $A \times A \rightarrow \MySet$, where $\Prop$ and $\MySet$ are types representing, that is classifying -- in category theory terminology --, propositions or sets, respectively. Along the indexed-fibered correspondence, that is along the equivalence between the types $A \rightarrow \Prop$ and $\Sigma B:\MySet. (B \stackrel{inj}{\rightarrow} A)$ (proof-irrelevant case) and $A \rightarrow \MySet$ and $\Sigma B:\MySet. (B \rightarrow A)$ (proof-relevant case), for $A:\MySet$, everything told in the indexed form can morally also be told in \emph{fibered} form. Such fibered approach to parametricity was for instance considered in Ghani \cite{Ghani...}, who also suggested to iterate parametricity.

@@

The internalisation of parametricity in unary indexed form was studied by \cite{Moulin,Bernardy} directly leading to a type theory of iterated parametricity. Such type theory could be interpreted in a ``unary'' form of cubical set~\cite{bernardy15}, thus implicitly in fibered form since, in a binary cubical set (resp. in a unary cubical set and simplicial set as well), a set of segments over two points, that is a relevant relation, is given by first and second projection rather than as a function to $\MySet$ (resp. a family is given as a \emph{total space} projected over a set of points $A:\MySet$ rather than a family of type $A \rightarrow \MySet$. This connection between iterated binary parametricity in fibered form and cubical sets includes an identification of degeneracies and reflexivity proofs, which can also be found in Johann and Sojakova~\cite{johann17}. In particular, it became clear at this point that iterated binary parametricity (without reflexivity) produces semi-cubical sets while it was also observed\footnote{The authors learned it from Hugo Moeneclaye and Thorsten Altenkirch.} that iterated unary parametricity produces augmented semi-simplicial sets.

$$
\begin{tabular}{lllllll}
                           &                            &                                 &                   & structure & kind of semantics \\
 \hline
 \multirow{6}{1cm}{unary}  & \multirow{3}{1.5cm}{indexed} & \multirow{2}{2cm}{non-iterated} & proof-irrelevant  & predicate &        reducibility \\
                           &                            &                                 & relevant          & family & \\
                           &                            & \multicolumn{2}{l}{iterated}                        & \emph{this work}       & \\
                           & \multirow{3}{1.5cm}{fibered} & \multirow{2}{2cm}{non-iterated} & proof-irrelevant  & subobject & \\
                           &                            &                                 & relevant          & fibration & \\
                           &                            & \multicolumn{2}{l}{iterated}                        & augmented semi-simplicial set & \\
 \hline
 \multirow{6}{1cm}{binary} & \multirow{3}{1.5cm}{indexed} & \multirow{2}{2cm}{non-iterated} & proof-irrelevant  & relation & logical relation \\
                           &                            &                                 & relevant          & correspondence (relevant relation) & \\
                           &                            & \multicolumn{2}{l}{iterated}                        & \emph{this work}       & \\
                           & \multirow{3}{1.5cm}{fibered} & \multirow{2}{2cm}{non-iterated} & proof-irrelevant  & fibered relation ? & \\
                           &                            &                                 & relevant          & span & & \\
                           &                            & \multicolumn{2}{l}{iterated}                        & semi-cubical set & \\
 \hline
\end{tabular}
$$

@@

Table~\ref{tab:coind} describes the type of a $\nu$-set at level $m$ as a dependent stream of type families representing the limit of $n$-truncated $\nu$-sets: the recursive equation $\Xfrom{m}{n}{D} \,\defeq\, \Sigma R: \Xcomp{m}{n}[D=\D].\, \Xfrom{m}{n+1}[D=\pair{D}{R}]$ from the table formally corresponds to the stream $Stream_{\Sigma n.\, \Xto{m}{n},\, \lambda (n, \D).\, \Xcomp{m}{n}[D=\D],\,\lambda ((n, D),R).(n+1,(\pair{D}{R}))}(n, \D)$. Therefore, $\Xfrom{m}{n}$ denotes an infinite sequence $X_{n}, X_{n+1}, \ldots$ dependent on a $(<n)$-truncated $\nu$-set, $\Xto{m}{n}$, so that, when $n$ is $0$, it denotes a full $\nu$-set, written $\Xp{m}$. This is made possible because the $(<0)$-truncated $\nu$-set, $\Xto{m}\lbrack 0,0 \rbrack$, is degenerated: it is an empty family, and there is thus only one $(<0)$-truncated $\nu$-set, namely the canonical inhabitant $\kstar$ of $\unittype$.

@@

Since $\restrf[n][p]$ is referring to $\framep[n][p]$ and $\framep[n-1][p]$, and $\cohf[n][p]$ is referring to $\restrf[n][p]$ and $\restrf[n-1][p]$, thus also to $\framep[n][p]$, $\framep[n-1][p]$ and $\framep[n-2][p]$, we built all of these 6 components at the same time. But equations were also needed to state the properties of $\framep[n][p]$, $\framep[n-1][p]$ and $\restrf[n][p]$ so that we were building by induction on $n$, for any $D:\Xto{m}{n}$, the following block of information:

\begin{equation*}
  \begin{array}{lclr}
    \framep[n][p]      & : & \U                                                                                                                 & \mbox{for all $p\leq n$}   \\
    \framep[n-1][p]    & : & \U                                                                                                                 & \mbox{for all $p\leq n-1$} \\
    \framep[n-2][p]    & : & \U                                                                                                                 & \mbox{for all $p\leq n-2$} \\
    \restrf[n][p]      & : & \Pi q.\, \Pi \omega.\, \framep[n][p] \rightarrow \framep[n-1][p]                                                   & \mbox{for all $p\leq n$}   \\
    \restrf[n-1][p]    & : & \Pi q.\, \Pi \omega.\, \framep[n-1][p] \rightarrow \framep[n-2][p]                                                 & \mbox{for all $p\leq n-1$} \\
    \cohf[n][p]        & : & \Pi qr\epsilon\omega.\, \restrf[n-1][p][q][\epsilon] \circ \restrf[n][p][\omega][r] =                                                           \\
                       &   & \restrf[n-1][p][r][\omega] \circ \restrf[n][p][\epsilon][q+1]                                                      & \mbox{for all $p\leq n$}   \\
    \mathsf{hypF0}     & : & \framep[n][0]         = \unitpoint
    \\
    \mathsf{hypFS}     & : & \framep[n][p+1]       = \Sigma d: \framep[n+1][p]. \Pi \omega.\, \painting[n][p] (\restrf[n+1][p](d))              & \mbox{for all $p\leq n-1$} \\
    \mathsf{hypPrevF0} & : & \framep[n-1][0]                                                                                       = \unitpoint                              \\
    \mathsf{hypPrevFS} & : & \framep[n-1][p+1] = \Sigma d: \framep[n][p]. \Pi \omega.\, \painting[n-1][p] (\restrf[n][p](d))                    & \mbox{for all $p\leq n-2$} \\
    \mathsf{hypRF0}    & : & \restrf[n][0]        = \ldots \texttt{depending on hypF0}                                                                                       \\
    \mathsf{hypRFS}    & : & \restrf[n][p+1]      = \ldots \texttt{depending on hypFS and hypPrevFS}                                            & \mbox{for all $p\leq n-1$} \\
    \ldots             &   & \texttt{fields for painting, restrf, cohp}
  \end{array}
\end{equation*}

In the current paper and associated formalisation, we switch the point
of view, take $\framep[n-2][p]$ $\painting[n-2][p]$, $\restrf[n-1][p]$
and $\restrp[n-1][p]$ as references and rename them respectively as
$\framep[n][p]$, $\painting[n][p]$, $\restrf[n][p]$ and
$\restrp[n][p]$. We replace the equational properties of $\framep[n][p]$,
$\framep[n-1][p]$, $\restrf[n][p]$, $\painting[n][p]$,
$\painting[n-1][p]$ and $\restrp[n][p]$ with definitions, renaming them
respectively to $\framep[n+2][p]$, $\framep[n+1][p]$,
$\restrf[n+1][p]$, $\painting[n+2][p]$, $\painting[n+1][p]$ and
$\restrp[n+1][p]$. This means that we build instead by induction a block of the form:

@@

              (\depscohs[n][\lbrack 0,p-1 \rbrack]),           \\

@@

      \restrl[n][p][q][\epsilon](\depscohs[n][\lbrack 0,p-1 \rbrack])(\restrl[n+1][p][r][\omega](\overline{\depscohs}(\depscoht[n][\lbrack 0,p-1 \rbrack]))(l))\\ =_{\cohf[n+1][p][r+1][q+1][\epsilon][\omega](\depscoht[n][\lbrack 0,p-1 \rbrack])(d)} \\

@@

      (\ap \overrightarrow{\cohf[n][p][r][0][\omega][\theta](\restrf[n+2][p][q+2][\epsilon](\overline{\depscohs}(\depscoht[n][\lbrack 0,p \rbrack]))(d))}\\ (\ap \overrightarrow{\cohf[n+1][p][q+1][0][\epsilon][\theta](\depscoht[n][\lbrack 0,p-1 \rbrack])(d)} \\(\cohp[n][p][q][r][\epsilon][\omega](l_\theta)))) \\

>>>>

Parametricity, like realisability, reducibility, and logical relations, was most commonly handled in an \emph{indexed} form, that is, using a type-theoretic language, considering predicates of type $A \rightarrow \Prop$, or families of type $A \rightarrow \Set$, or relations of type $A \times A \rightarrow \Prop$, or relevant relations -- sometimes called \emph{correspondences} -- of type $A \times A \rightarrow \Set$, where $\Prop$ and $\Set$ are types representing, that is classifying -- in category theory terminology --, propositions or sets, respectively. Along the indexed-fibered correspondence, that is along the equivalence between the types $A \rightarrow \Prop$ and $\Sigma B:\Set. (B \stackrel{inj}{\rightarrow} A)$ (proof-irrelevant case) and $A \rightarrow \Set$ and $\Sigma B:\Set. (B \rightarrow A)$ (proof-relevant case), for $A:\Set$, everything told in the indexed form can morally also be told in \emph{fibered} form. Such fibered approach to parametricity was for instance considered in Ghani \cite{Ghani...}, who also suggested to iterate parametricity.

@@

The internalisation of parametricity in unary indexed form was studied by \cite{Moulin,Bernardy} directly leading to a type theory of iterated parametricity. Such type theory could be interpreted in a ``unary'' form of cubical set~\cite{bernardy15}, thus implicitly in fibered form since, in a binary cubical set (resp. in a unary cubical set and simplicial set as well), a set of segments over two points, that is a relevant relation, is given by first and second projection rather than as a function to $\Set$ (resp. a family is given as a \emph{total space} projected over a set of points $A:\Set$ rather than a family of type $A \rightarrow \Set$. This connection between iterated binary parametricity in fibered form and cubical sets includes an identification of degeneracies and reflexivity proofs, which can also be found in \cite{johann17}. In particular, it became clear at this point that iterated binary parametricity (without reflexivity) produces semi-cubical sets while it was also observed\footnote{The authors learned it from Hugo Moeneclaye and Thorsten Altenkirch.} that iterated unary parametricity produces augmented semi-simplicial sets.

\begin{tabular}{lllllll}
                            &                              &                                 &                               & structure                          & kind of semantics   \\
  \hline
  \multirow{6}{1cm}{unary}  & \multirow{3}{1.5cm}{indexed} & \multirow{2}{2cm}{non-iterated} & proof-irrelevant              & predicate                          & reducibility        \\
                            &                              &                                 & relevant                      & family                             &                     \\
                            &                              & \multicolumn{2}{l}{iterated}    & \emph{this work}              &                                                          \\
                            & \multirow{3}{1.5cm}{fibered} & \multirow{2}{2cm}{non-iterated} & proof-irrelevant              & subobject                          &                     \\
                            &                              &                                 & relevant                      & fibration                          &                     \\
                            &                              & \multicolumn{2}{l}{iterated}    & augmented semi-simplicial set &                                                          \\
  \hline
  \multirow{6}{1cm}{binary} & \multirow{3}{1.5cm}{indexed} & \multirow{2}{2cm}{non-iterated} & proof-irrelevant              & relation                           & logical relation    \\
                            &                              &                                 & relevant                      & correspondence (relevant relation) &                     \\
                            &                              & \multicolumn{2}{l}{iterated}    & \emph{this work}              &                                                          \\
                            & \multirow{3}{1.5cm}{fibered} & \multirow{2}{2cm}{non-iterated} & proof-irrelevant              & fibered relation ?                 &                     \\
                            &                              &                                 & relevant                      & span                               &                   & \\
                            &                              & \multicolumn{2}{l}{iterated}    & semi-cubical set              &                                                          \\
  \hline
\end{tabular}

@@

Table~\ref{tab:coind} describes the type of a $\nu$-set at level $m$ as a dependent stream of type families representing the limit of $n$-truncated $\nu$-sets: the recursive equation $\Xfrom{m}{n}{D} \,\defeq\, \Sigma R: \Xcomp{m}{n}[D=\D].\, \Xfrom{m}{n+1}[D=\pair{D}{R}]$ from the table formally corresponds to the stream $Stream_{\Sigma n.\, \Xto{m}{n},\, \lambda (n, \D).\, \Xcomp{m}{n}[D=\D],\,\lambda ((n, D),R).(n+1,(\pair{D}{R}))}(n, \D)$. Therefore, $\Xfrom{m}{n}$ denotes an infinite sequence $X_{n}, X_{n+1}, \ldots$ dependent on a $(<n)$-truncated $\nu$-set, $\Xto{m}{n}$, so that, when $n$ is $0$, it denotes a full $\nu$-set, written $\Xp{m}$. This is made possible because the $(<0)$-truncated $\nu$-set, $\Xto{m}{0}$, is degenerated: it is an empty family, and there is thus only one $(<0)$-truncated $\nu$-set, namely the canonical inhabitant $\kstar$ of $\unittype$.

@@

In the current paper and associated formalisation, we take $\framep[n-2][p]$ $\painting[n-2][p]$, $\restrf[n-1][p]$ and $\restrp[n-1][p]$ as references and rename them respectively as $\framep[n][p]$, $\painting[n][p]$, $\restrf[n][p]$ and $\restrp[n][p]$. We replace the equational properties of $\framep[n][p]$, $\framep[n-1][p]$, $\restrf[n][p]$, $\painting[n][p]$, $\painting[n-1][p]$ and $\restrp[n][p]$ with definitions, renaming them respectively to $\framep[n+2][p]$, $\framep[n+1][p]$, $\restrf[n+1][p]$, $\painting[n+2][p]$, $\painting[n+1][p]$ and $\restrp[n+1][p]$. This means that we build instead by induction a block of the form:

@@

              (\depscohs[n][\lbrack 0,p-1 \rbrack]),

@@

      \restrl[n][p][q][\epsilon](\depscohs[n][\lbrack 0,p-1 \rbrack])(\restrl[n+1][p][r][\omega](\overline{\depscohs}(\depscoht[n][\lbrack 0,p-1 \rbrack]))(l)) \\ =_{\cohf[n+1][p][r+1][q+1][\epsilon][\omega](\depscoht[n][\lbrack 0,p-1 \rbrack])(d)} \\

@@

      (\ap \overrightarrow{\cohf[n][p][r][0][\omega][\theta](\restrf[n+2][p][q+2][\epsilon](\overline{\depscohs}(\depscoht[n][\lbrack 0,p \rbrack]))(d))} \\ (\ap \overrightarrow{\cohf[n+1][p][q+1][0][\epsilon][\theta](\depscoht[n][\lbrack 0,p-1 \rbrack])(d)} \\(\cohp[n][p][q][r][\epsilon][\omega](l_\theta)))) \\
